### 2024-03-29 parts/intro.tex
rewrite intro

\AP To approach the \emph{aperiodicity} membership problems the traditional
approach has been to develop so-called canonical objects, akin to the minimal
automaton or the syntactic monoid of regular languages, over which aperiodicity
is equivalent to a simple syntactic criterion (e.g. having no counters). This
is the proof scheme that has been implemented for sequential functions
\cite{CHOF03}, and more recently for rational functions \cite{FGL18}.
Unfortunately, such a canonical object remains to be defined for regular and
polyregular functions.

@@

\AP It was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper. Arrows denote strict inclusions,
and effectiveness (both in terms of decidability and of effective
representation) is represented by thick double arrows. Inclusions that are
suspected to be effective are represented using a dashed arrow together with a
question mark.

@@

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
$\ZPoly$ are all decidable, with effective conversion algorithms. This
effectively shows that all arrows in the lower left square of
\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
function} as the class of \kl{$\Rel$-polyregular functions} that are
\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
As a byproduct of our analysis, we provide a counter example to the
characterization of polynomials computed by \kl{$\Nat$-rational series} given
by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

@@

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture} holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

>>>>

\AP To approach the \emph{aperiodicity} membership problems the traditional
approach has been to develop so-called canonical objects, akin to the minimal
automaton or the syntactic monoid of regular languages, over which aperiodicity
is equivalent to a simple syntactic criterion (e.g. having no counters). This
is the proof scheme that has been implemented for sequential functions
\cite{CHOF03}, and more recently for rational functions \cite{FGL16,FGLM18}.
Unfortunately, such a canonical object remains to be defined for regular and
polyregular functions.

@@

\AP It was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold: $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper. Arrows denote strict inclusions,
and effectiveness (both in terms of decidability and of effective
representation) is represented by thick double arrows. Inclusions that are
suspected to be effective are represented using a dashed arrow together with a
question mark.

@@

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cite[Conjecture
7.61]{DOUE23} (restated in \cref{zsf-nsf:conjecture}), and prove that the
membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
decidable, with effective conversion algorithms. This effectively shows that
all arrows in the upper left square of \cref{previously-known-inclusions:fig}
are decidable in the \kl{commutative} setting. Furthermore, we characterize
\kl{commutative} \kl{$\Nat$-polyregular function} as the class of
\kl{$\Rel$-polyregular functions} that are \kl{combinatorial} (defined page
\kpageref{combinatorial}), which is a semantic property akin to the one
\kl{ultimate polynomiality} that characterizes $\ZSF$ inside $\ZPoly$
\cite{CDTL23}. As a byproduct of our analysis, we provide a counter example to
the characterization of polynomials computed by \kl{$\Nat$-rational series}
given by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

@@

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture}
\cite[Conjecture 7.61]{DOUE23} 
holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

### 2024-03-29 parts/prelim.tex
rewrite intro

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{THOM97}).

\AP In order to build functions in $\NPoly$, the following constructions will
be useful. For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon
\Sigma^* \to \set{0,1}$ as the indicator function of $L$. Given an $\MSO$
(resp. $\FO$) formula with first-order free variables $\varphi(\vec{x})$, we
define $\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$
that maps a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$,
such that $w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from
$\Sigma^*$ to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
multiplication with $s \in \mathbb{S}$.

@@

\AP
% How the problem has been approached so far
To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{BOJA22}.

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}. Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
x^n$.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a production function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
    \proofref{size-of-word-nsf:ex}
\end{example}

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{DOUE23,CDTL23}.

@@

\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
limiting the maximal number of free variables used to construct the function as
a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
$k$ free variables. It has been proven in
\cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
\bigO\left(\card{w}^k\right)$, and similarly for the other classes.

\subsection{Commutative Input}

\AP 
Let $\Sigma$ be a finite alphabet. 
We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
map that counts occurrences of every letter in the input word (that is,
computes the Parikh vector) namely: $ \commute[w] \defined (a \mapsto
\card[a]{w})$. Let $X$ be a set, a function $f \colon \Sigma^* \to X$ is
\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
\Nat^\Sigma \to X$ such that $g \circ \commute = f$.

To ensure that we are working inside a decidable subclass of \kl{$\Rel$-rational series}
let us state that commutativity is decidable. The proof is straightforward in the case
of \kl{$\Rel$-polyregular functions}, and more involved for general \kl{$\Rel$-rational series}.

@@

\AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function
$f \colon \Sigma^* \to \Rel$ if there exists a map $\eta \colon \set{1, \dots,
n} \to \Sigma$ such that $g \circ (v \mapsto \seqof{v_{\eta(k)}}{1 \leq k \leq
n}) \circ \commute = f$. This notion will be useful to formally state that a
polynomial ``is'' a \kl{polyregular function}.

>>>>

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{THOM97}). Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
x^n$.

\AP We also assume the reader to be familiar with rational series and weighted
automata, and refer to \cite{BERE10} for a comprehensive introduction. Let us
recall that one of the equivalent definitions of a $K$-rational series (where
$K$ is a ring) is given by weighted automata, which are non-deterministic
finite automata with transitions labelled by elements of $K$, and whose output
on a given word is the sum over all accepting paths of the products of the
weights along the path.

\AP We use the notation $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the map
that counts occurrences of every letter in the input word (that is, computes
the Parikh vector) namely: $ \commute[w] \defined (a \mapsto \card[a]{w})$.
Given a set $X$, a function $f \colon \Sigma^* \to X$ is \intro{commutative}
whenever for all $u \in \Sigma^*$, for all permutations $\sigma$ of $\set{1,
\dots, \card{w}}$, $f(\sigma(u)) = f(u)$. Equivalently, it is
\reintro{commutative} whenever there exists a map $g \colon \Nat^\Sigma \to X$
such that $g \circ \commute = f$. A function $g \colon \Nat^n \to X$ is
\intro{represented} by a function $f \colon \Sigma^* \to X$ if there exists a
map $\eta \colon \set{1, \dots, n} \to \Sigma$ such that $g \circ (v \colon
\Nat^\Sigma \mapsto \seqof{v(\eta(k))}{1 \leq k \leq n} \colon \Nat^n) \circ
\commute = f$. This notion will be useful to formally state that a polynomial
``is'' a \kl{polyregular function}.

@@

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
\end{example}
\begin{proof}
    Let us define $M \defined (\set{1}, \times)$ which is 
    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
    defined by $\mu(w) \defined 1$, and
    $\pi \colon M^2 \to \Nat$
    that is the constant function equal to $1$.
    We check that for all $w \in \Sigma^*$:
    \begin{equation*}
        \pi^\dagger(w)
        =
        \sum_{uv = w} 1
        =
        \card{w} + 1
        = f(w)
        \quad . 
        \qedhere
    \end{equation*}
\end{proof}

\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
defined as the minimal $k$ such that $\card{f(w)} =
\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp. $\reintro*\NPoly[k]$) for
the functions in $\ZPoly$ (resp. $\NPoly$) that have growth rate at most $k$.
It is clear that a function $f \in \ZPoly$ defined using a production function
$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
generally it is known that \kl{polyregular functions} have \kl{polynomial
growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
growth}, a typical example being the function $a^n \mapsto 2^n$.

\AP Let us briefly recall in the following lemmas alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
first-order free variables $\varphi(\vec{x})$, we define
$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
multiplication with elements $s \in \mathbb{S}$.

@@

It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
of first order free variables used to construct the function as in
\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.

Let us briefly state that \kl{commutativity} is a decidable property of
\kl{$\Rel$-rational series}. The proof is straightforward in the case of
\kl{$\Rel$-polyregular functions}, and slightly more involved for general
\kl{$\Rel$-rational series}.

@@
